### 2018-11-20 Manuscript/Manuscript.tex

One of the hypothesized effects of the emplacement of large igneous provinces (LIPs) is planetary cooling on million-year timescales associated with enhanced silicate weathering. A systematic analysis of LIP area and latitude through the Phanerozoic reveals that there is no significant correlation between total LIP area, nor LIP area in the tropics, and the extent of continental ice sheets. The largest peaks in tropical LIP area are at times of non-glacial climate. These results suggest that changes in planetary weatherability associated with LIPs are not the fundamental control on whether Earth is in a glacial or non-glacial climate, although they could provide a secondary modulating effect in conjunction with other processes.

@@

Global weatherability is the sum of factors aside from climate itself, such as the latitudinal distribution of continents and mountain belts, that contribute to overall global weathering and associated CO$_2$ consumption. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. Basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust because of the relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff, and as a result CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2009a, Hartmann2014a}.

One aspect of large igneous province (LIP) emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. In particular, the emplacement of LIPs in the tropics has been hypothesized to be associated with specific episodes of climatic cooling on Earth. In the Neoproterozoic Era, the emplacement of the ca. 720~Ma Franklin LIP in the tropics, in concert with elevated runoff rates associated with continental break-up, has been implicated as a major contributor to the cooling that initiated the Sturtian `Snowball Earth' \citep{Donnadieu2004b, Macdonald2010a, Cox2016a}. In the Cenozoic, the movement of the Deccan LIP into the tropical rain belt, in concert with the low-latitude emplacement of the Ethiopian Traps, has been implicated in drawing down CO$_2$ levels in the lead-up to Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}.

@@

Outlines of the original surface extent of LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were slightly modified from the compilation of \citet{Ernst2017a}, and emplacement ages were taken from the literature (Table \ref{tab:LIPs}). The compilation of \citet{Ernst2017a} seeks to reconstruct the original surface extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP, including dikes, sills, and layered intrusions in addition to subaerial volcanics (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original surface extent, given that subsurface intrusions could extend over a broader area than the surface volcanics. The polygons also assume complete surface coverage between wide-spread remnants, creating further potential for these outlines being over-estimates. However, this approach likely provides the best estimates available of original extent for ancient LIPs. The extents of presently-exposed volcanics associated with LIPs that were used for present-day area estimates, and the construction of the original extent polygons by \citet{Ernst2017a}, were taken from a number of resources including the PLATES compilation \citep{Coffin2006a} and more recent compilation efforts associated with the LIPs Reconstruction Project (\citealp{Ernst2013a}; Table \ref{tab:LIPs}).

After LIPs are emplaced, they will progressively erode. In order to account for this decrease in area with time, \citet{Godderis2017a} took the approach of fitting an exponential decay function to estimates of the original surface extent and the current surface extent for 5 LIPs. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original surface extent of the province and the current surface extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate first-order model for the progressive reduction in LIP area. The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude from our estimation of a representative LIP area half-life the 6 of 19 LIPs which have been partially or completely buried. This latter approach yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). Although the root mean square error of the fit is low (0.14), there is variability in the half-lives for LIPs if they are fit individually with an exponential decay function, such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps (Table \ref{tab:LIPs}). In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36~Myr' uses the best-fit half-life of 36~Myr, while `t$_{1/2}$ = 120~Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstructions used in this study also include pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly impact the calculated LIP areas throughout the Phanerozoic, which is the focus of this analysis.

@@

Our LIP reconstruction differs from that of \citet{Johansson2018a} in several respects. In contrast to the decay and decay+burial scenarios implemented on estimates of original LIP extent in this study, \citet{Johansson2018a} uses a static extent for each LIP throughout the reconstruction with some of the polygons corresponding to the present-day surface extent and some representing the original extent that includes currently buried portions of the LIP (e.g. the Keweenawan Midcontinent Rift for which the implemented extent is from geophysical data that largely correponds to buried subsurface exposures).

@@

In the `t$_{1/2}$ = 36~Myr' scenario, we observe three primary peaks and one minor peak in the calculated LIP area within the tropics (Fig. \ref{fig:LIP_area}C). The first primary peak ca. 510~Ma is associated with the emplacement of the Kalkarindji LIP, the second primary peak ca. 380~Ma is associated with the emplacement of the Kola-Dnieper LIP, the third primary peak ca. 200~Ma is associated with the emplacement of the CAMP, and the minor Cenozoic peak is associated with both the ca. 30~Ma emplacement of the Afar LIP as well as the earlier drift of the ca. 66~Ma Deccan LIP into the tropics (Fig. \ref{fig:LIP_area}A). When we account for burial (`t$_{1/2}$ = 36~Myr + 50\% burial' and `t$_{1/2}$ = 36~Myr + 100\% burial' scenarios), only the latter two of these four peaks are affected -- the ca. 200~Ma peak is attenuated/removed due to the partial/complete burial of the CAMP, and the Cenozoic peak is attenuated due to the partial/complete removal of the Afar LIP. However, after accounting for burial, a minor area of LIPs remain in the tropics from ca. 130~Ma onwards, due to the EQUAMP, Caribbean-Colombian, and Deccan LIPs. Using the longer decay half-life of 120~Myr (the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario) increases the area of LIPs in the tropics at any given time step, and has the effect of extending the duration of each peak.

The only scenario which results in a positive Pearson correlation coefficient (0.10) between LIP area in the tropics (Fig. \ref{fig:LIP_area}C) and the ice extent record (Fig. \ref{fig:LIP_area}E) is the scenario with the slow decay rate and complete burial of LIPs associated with rifting (the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario; Fig. \ref{fig:LIP_correlation}). All other scenarios yield a near zero or weak negative correlation coefficient. Qualitatively, the weak positive correlation of the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario relative to the other scenarios can be primarily attributed to the complete removal of the CAMP, which was emplaced during an extended interval of ice-free conditions, as well as the effect of the longer decay half-life extending the duration of the earlier two peaks, such that they overlap more with the Late Ordovician and Permo-Carboniferous glacial intervals.

@@

@@

In the original models that proposed the `Fire and Ice' hypothesis as an explanation for the onset of the Sturtian `Snowball Earth' glaciation, chemical weathering was modeled as an Arrhenius relationship where it was solely a function of temperature and run-off \citep{Donnadieu2004a}. However, such an approach neglects the effects of soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography, and chemical weathering reveals that these effects are quite important \citep{Gabet2009a, Maher2014a}. As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones \citep{Godderis2017b}. Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisions in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$520~Myr. Furthermore, the approach of seeking to evaluate correlation between LIP area and glaciation is complicated for Neoproterozoic Snowball Earth events, since ice-albedo runaway leads to persistent global glaciation absent of continued forcing through normal carbon cycle processes until sufficient CO$_2$ builds up in the absence of silicate weathering to drive deglaciation. Nevertheless, the hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth is a major motivating driver behind conducting this analysis. How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? Using the paleomagnetic pole of \citet{Denyszyn2009a}, we reconstruct the paleolatitude of the Franklin LIP at the time of emplacement, and find that $\sim$99.7\% (or $\sim$2.6~Mkm$^{2}$) of the LIP erupted within 15\textdegree\xspace of the equator. This Franklin LIP tropical area at the time of emplacement is approximately equivalent to the Cenozoic peak, and is smaller than the other Phanerozoic peaks (Fig. \ref{fig:LIP_area}C). The ca. 1109~Ma Umkondo is another Precambrian LIP that is constrained to have erupted in the tropics, although it is not known to be associated with any glaciation (no glacial deposits are found within the contemporaneous Midcontinent Rift basin; \citealp{Swanson-Hysell2018a}). We reconstruct the paleolatitude of the Umkondo LIP at the time of emplacement using the paleomagnetic pole of \citet{Swanson-Hysell2015b}, and find that effectively all of the LIP (or $\sim$2.0~Mkm$^{2}$) erupted within the tropics, an area that is slightly smaller than that for the Franklin LIP (Fig. \ref{fig:LIP_area}C). Together, these results indicate that the tropical area of the Franklin LIP, when compared to that of Phanerozoic as well as other Precambrian LIPs, did not have a singularly large area in the tropics. Given that similar peaks in tropical LIP area are not associated with the onset of glacial periods, additional processes must have been at play in drawing down atmospheric CO$_2$ in the lead-up to the Sturtian Snowball Earth. One such process could have been an increase in global weatherability associated with arc accretion within terranes of the present-day Arabian-Nubian Shield \citep{Park2018a}. An increase in global weatherability associated with the emplacement of the Franklin LIP in the tropics could have been a contributing factor in addition to other such processes. The temporal overlap between Franklin LIP eruptions \citep{Denyszyn2009a} and the initiation of Sturtian glaciation \citep{Macdonald2010a, MacLennan2018a} remains compelling. This overlap continues to suggest that other aspects of LIP emplacement, such as the injection of sulfur aerosols in the stratosphere \citep{Macdonald2017a}, could have played a role in the initiation of low-latitude glaciation. The latitude of such aerosol injections is important for their temporary effect on albedo which is maximized at low latitude, and their presence at high concentrations is pre-conditioned on eruption through sedimentary basins hosting evaporite deposits \citep{Macdonald2017a}. Overall, the conclusion that tropical LIP area associated with the Franklin is not uniquely high and therefore not a sole driver of Snowball Earth onset is consistent with the results of the Phanerozoic analysis. LIP area and its associated impact on global weatherability is not the fundamental control on Earth's long-term climate state.

@@

Richard Ernst provided GIS compilations of LIP extent and present-day exposure that were essential to the analysis. Discussions with Yves Godd\'eris were valuable in contributing to aspects of the interpretation. Park was supported by NSF Grant 1547434 awarded to Swanson-Hysell.

>>>>

One of the hypothesized effects of the emplacement of large igneous provinces (LIPs) is planetary cooling on million-year timescales associated with enhanced silicate weathering. This study combines outlines of the original surface extent of LIPs through the Phanerozoic, a paleogeographic model, and LIP erosion models to estimate LIP area in all latitudinal bands through the Phanerozoic. This analysis reveals no significant correlation between total LIP area, nor LIP area in the tropics, and the extent of continental ice sheets. The largest peaks in tropical LIP area are at times of non-glacial climate. These results suggest that changes in planetary weatherability associated with LIPs are not the fundamental control on whether Earth is in a glacial or non-glacial climate, although they could provide a secondary modulating effect in conjunction with other processes.

@@

Global weatherability is the sum of factors aside from climate itself, such as the latitudinal distribution of continents and mountain belts, that contribute to overall global weathering and associated CO$_2$ consumption. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. Basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust because of the relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff. As a result, CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2009a, Hartmann2014a}.

One aspect of large igneous province (LIP) emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. In particular, the emplacement of LIPs in the tropics has been hypothesized to be associated with specific episodes of climatic cooling on Earth. In the Neoproterozoic Era, the emplacement of the ca. 720~Ma Franklin LIP in the tropics, in concert with elevated runoff rates associated with continental break-up, has been implicated as a major contributor to the cooling that initiated the Sturtian `Snowball Earth' \citep{Donnadieu2004b, Macdonald2010a, Cox2016a}. In the Cenozoic, the movement of the Deccan LIP into the tropical rain belt, in concert with the low-latitude emplacement of the Ethiopian Traps, has been implicated in drawing down CO$_2$ levels in the lead-up to Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. More recently, \citet{Johansson2018a} used a paleogeographic model to suggest that tropical LIP area correlates with Phanerozoic climate change.

@@

This study combines outlines of the original surface extent of LIPs through the Phanerozoic, a paleogeographic model, and LIP erosion models to estimate LIP area in all latitudinal bands through the Phanerozoic. We then compare these LIP area time series to the latitudinal extent of continental ice sheets - a proxy for Earth's long term climate state.

Outlines of the original surface extent of LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were slightly modified from the compilation of \citet{Ernst2017a}, and emplacement ages were taken from the literature (Table \ref{tab:LIPs}). The compilation of \citet{Ernst2017a} seeks to reconstruct the original surface extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP, including dikes, sills, and layered intrusions, in addition to subaerial volcanics (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original surface extent, given that subsurface intrusions could extend over a broader area than the surface volcanics. The polygons also assume complete surface coverage between wide-spread remnants, creating further potential for these outlines being over-estimates. However, this approach likely provides the best estimates available of original extent for ancient LIPs. The extents of presently-exposed volcanics associated with LIPs that were used for present-day area estimates, and the construction of the original extent polygons by \citet{Ernst2017a}, were taken from a number of resources including the PLATES compilation \citep{Coffin2006a} and more recent compilation efforts associated with the LIPs Reconstruction Project (\citealp{Ernst2013a}; Table \ref{tab:LIPs}).

After LIPs are emplaced, they will progressively erode. In order to account for this decrease in area with time, \citet{Godderis2017a} took the approach of fitting an exponential decay function to estimates of the original surface extent and the current surface extent for 5 LIPs. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original surface extent of the province and the current surface extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate first-order model for the progressive reduction in LIP area (Fig. \ref{fig:LIP_preservation}). The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude from our estimation of a representative LIP area half-life the 6 of 19 LIPs which have been partially or completely buried. This latter approach yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). Although the exponential fit to the 13 unburied LIPs is good (i.e. it yields a low root mean square error of 0.14), if each LIP is fit individually with an exponential decay function, there is variability in the estimated half-lives such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps (Table \ref{tab:LIPs}). In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36~Myr' uses the best-fit half-life of 36~Myr, while `t$_{1/2}$ = 120~Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstructions used in this study also include pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly influence the calculated LIP areas throughout the Phanerozoic, which is the focus of this analysis.

@@

Our LIP reconstruction differs from that of \citet{Johansson2018a} in several respects. In contrast to the decay and decay+burial scenarios implemented on estimates of original LIP extent in this study, \citet{Johansson2018a} uses a static extent for each LIP throughout the reconstruction. In their analysis, some of the polygons correspond to the present-day surface extent and some represent the original extent that includes currently buried portions of the LIP (e.g. the Keweenawan Midcontinent Rift for which the implemented extent is from geophysical data that largely corresponds to buried subsurface exposures).

@@

In the `t$_{1/2}$ = 36~Myr' scenario, we observe three major peaks and one minor peak in the calculated LIP area within the tropics (Fig. \ref{fig:LIP_area}C). The first major peak ca. 510~Ma is associated with the emplacement of the Kalkarindji LIP, the second major peak ca. 380~Ma is associated with the emplacement of the Kola-Dnieper LIP, the third major peak ca. 200~Ma is associated with the emplacement of the CAMP, and the minor Cenozoic peak is associated with both the ca. 30~Ma emplacement of the Afar LIP as well as the earlier drift of the ca. 66~Ma Deccan LIP into the tropics (Fig. \ref{fig:LIP_area}A). When we account for burial (`t$_{1/2}$ = 36~Myr + 50\% burial' and `t$_{1/2}$ = 36~Myr + 100\% burial' scenarios), only the latter two of these four peaks are affected -- the ca. 200~Ma peak is attenuated/removed due to the partial/complete burial of the CAMP, and the Cenozoic peak is attenuated due to the partial/complete removal of the Afar LIP. However, after accounting for burial, a minor area of LIPs remain in the tropics from ca. 130~Ma onwards, due to the EQUAMP, Caribbean-Colombian, and Deccan LIPs. Using the longer decay half-life of 120~Myr (the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario) increases the area of LIPs in the tropics at any given time step, and has the effect of extending the duration of each peak.

The only scenario which results in a positive Pearson correlation coefficient (0.10) between LIP area in the tropics (Fig. \ref{fig:LIP_area}C) and the ice extent record (Fig. \ref{fig:LIP_area}E) is the scenario with the slow decay rate and complete burial of LIPs associated with rifting (the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario; Fig. \ref{fig:LIP_correlation}). All other scenarios yield a near zero or weak negative correlation coefficient. The weak positive correlation of the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario relative to the other scenarios can be primarily attributed to the complete removal of the CAMP, which was emplaced during an extended interval of ice-free conditions, as well as the effect of the longer decay half-life extending the duration of the earlier two peaks, such that they overlap more with the Late Ordovician and Permo-Carboniferous glacial intervals.

@@

In the original models that proposed the `Fire and Ice' hypothesis as an explanation for the onset of the Sturtian `Snowball Earth' glaciation, chemical weathering was modeled as a function of temperature and run-off only \citep{Donnadieu2004a}. However, such an approach neglects the effects of soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography, and chemical weathering reveals that these effects are quite important \citep{Gabet2009a, Maher2014a}. As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones \citep{Godderis2017b}. Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisions in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$520~Myr. The approach of seeking to evaluate correlation between LIP area and glaciation is further complicated for Neoproterozoic Snowball Earth events because ice-albedo runaway leads to persistent global glaciation on timescales of tens of millions of years absent of continued forcing through normal carbon cycle processes until sufficient CO$_2$ to drive deglaciation builds up in the absence of silicate weathering \citep{Hoffman2017a}. Moreover, cooling past the critical threshold for rapid global glaciation may have occurred on a sub-million year timescale that is not recorded in the geological carbon cycle (e.g. \citealp{Macdonald2017a}). Nevertheless, the hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth is a major motivating driver behind conducting this analysis.

How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? Using the paleomagnetic pole of \citet{Denyszyn2009a}, we reconstruct the paleolatitude of the Franklin LIP at the time of emplacement, and find that $\sim$99.7\% (or $\sim$2.6~Mkm$^{2}$) of the LIP erupted within 15\textdegree\xspace of the equator. This Franklin LIP tropical area at the time of emplacement is approximately equivalent to the Cenozoic peak, and is smaller than the other Phanerozoic peaks (Fig. \ref{fig:LIP_area}C). The ca. 1109~Ma Umkondo is another Precambrian LIP that is constrained to have erupted in the tropics, although it is not known to be associated with any glaciation (no glacial deposits are found within the contemporaneous Midcontinent Rift basin; \citealp{Swanson-Hysell2018a}). We reconstruct the paleolatitude of the Umkondo LIP at the time of emplacement using the paleomagnetic pole of \citet{Swanson-Hysell2015b}, and find that effectively all of the LIP (or $\sim$2.0~Mkm$^{2}$) erupted within the tropics, an area that is slightly smaller than that for the Franklin LIP (Fig. \ref{fig:LIP_area}C).

Together, these results indicate that the tropical area of the Franklin LIP, when compared to that of Phanerozoic as well as other Precambrian LIPs, did not have a singularly large area in the tropics. Given that similar peaks in tropical LIP area are not associated with the onset of glacial periods, additional processes beyond an increase in weatherability due to LIP area in the tropics must have been at play in the initiation of the Sturtian Snowball Earth. One such process could have been an increase in global weatherability associated with the collision and accretion of arc terranes within the present-day Arabian-Nubian Shield \citep{Park2018a}. A low-latitude continent configuration and abundant arc-continent collisions in the tropics may have led to a cool background climate and the emplacement of the Franklin LIP may have pushed the climate system over the edge. However, tropical LIP area associated with the Franklin was not uniquely high, and therefore an associated increase in global weatherability was not the sole driver of Snowball Earth onset, consistent with the results of the Phanerozoic analysis.

An increase in global weatherability associated with the emplacement of the Franklin LIP in the tropics could have been a contributing factor in addition to other such processes. The temporal overlap between Franklin LIP eruptions and the initiation of Sturtian glaciation remains compelling \citep{Macdonald2017a}. This overlap continues to suggest that other aspects of LIP emplacement, such as relieving P, Fe, or S limitations on biological productivity \citep{Tziperman2011a, Cox2016a}, or the injection of sulfur aerosols in the stratosphere \citep{Macdonald2017a}, could have played a role in the initiation of low-latitude glaciation. The latitude and background climate of such aerosol injections is important for their temporary effect on albedo which is maximized at low latitude, and their presence at high concentrations is pre-conditioned on eruption through sedimentary basins hosting evaporite deposits \citep{Macdonald2017a}. Sulfur aerosols associated with low-latitude LIPs may have also caused short-term cooling due to changes in albedo during the Phanerozoic, as has been suggested for the CAMP \citep{Schoene2010a} and putative LIPs associated with the Hirnantian ice advance \citep{Jones2017a}. However, because these cooling events did not result in a Snowball Earth, the climate would have returned to its background climate state within thousands of years \citep{Macdonald2017a}, which would not be observed in our analysis.

@@

@@

Richard Ernst provided GIS compilations of LIP extent and present-day exposure that were essential to the analysis. Discussions with Yves Godd\'eris made possible through the France Berkeley Fund were valuable in contributing to aspects of the interpretation. Park was supported by NSF Grant 1547434 awarded to Swanson-Hysell.
